VF1 Vignette

### 2024-03-12 vignette.Rmd
Commit before restructuring.

# What is visual field testing?
Visual field testing is a psychophysical test that measures a patient’s visual ability at different locations of the visual field. This ability is usually highest in the center of the visual field (i.e. the point that the observer looks at) and decreases toward the periphery (hill of vision).

@@

```
This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has a chronic-neurodegenerative disease of the optic nerve called primary open-angle glaucoma (type: pwg).

@@

# Visual field tests
For the typical visual field test, the subject or patient is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld cupule (or hemisphere). The perimeter is located in a dimly lit, calm room and the subject has been explained what to do by an experienced technician before the examination. In his hand, the subject has a button in order to signal whether he saw a stimulus.

@@

```
The standard environment assumes a static white-on-white perimetry with a Humphrey perimeter, using a 24-2 grid, a SITA standard algorithm, a stimulus size of Goldman Size III. The stimulus duration and the $L_{max}$ are not explicitly stated in the info.

@@

The spatial position of the points $l_1 $to $ln$ in the visual field is found in the locmap and depends on the pattern used by the perimeter. When analysing data, you need to specify the locmap.

@@

Note that some grids spare the blind spot, either on purpose or because they cover only locations central to the blind spot. In this case, locmap$bs is numeric(0).

@@

# Pathophysiology of the visual field
Interindividual variability and age
Photoreceptors / Retina
Ganglion cells
Other factors

# Measurements
Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response. It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.

@@

# Differential light sensitivity and other mathematical derivatives of threshold
However, the data analyzed with the visual fields package usually consists of the subjects *differential light sensitivity* for a stimulus at a given location in decibel (dB) as the most elementary unit. This sensitivity is calculated from the psychophysical threshold, which is derived from the aforementioned responses at this location.

@@

## False negatives / positives
Residual deviation from the psychometric function may occur due to inattention of the observer but also due to short-term fluctuation of the threshold. For example, it has been shown that the threshold underlies short-term fluctuates in the areas where light sensitivity was negatively affected by glaucoma, a disease of the optic nerve.

@@

The resulting difference is also measured in dB and is called either *deviaton* (loss of function has a negative sign; used in Humphrey perimeters) or *defect* (loss of function has a positive sign; used in Octopus perimeters). In this package, *deviation* is used, and it is called *total deviation*. Make sure that you get the right sign if you import data.

@@

### Sensitivity

@@

vfpwgSunyiu24d2[1:4, "1:12"]

@@

@@

The standard already contains some additional information with a border presenting a color-coded visualisation of the deviation and a graphical representation of the blind spot.

@@

@@

```
Again, the probability of the observed values given that the visual field is normal can be calculated.

@@

Monitoring function over time in glaucoma patients is the most common clinical application of visual fields in ophthalmology. For the next example, we will select a series of 27 visual fields that were obtained over several years.

@@

@@

vflegoplot(example2)
vfsparklines(example2)

@@

@@

@@

```
```{r}

res <- plr(example2)

@@

@@

poplr(example2)

@@

@@

vfspa(example2, file = "test.pdf")
# vfspashiny(example2, file = "test.pdf")

@@

DICOM (Digital Imaging and Communications in Medicine) standards for ophthalmic static perimetry are specified in Supplement 146. Currently, the data representation in the VisualFields package differs from these standards. However, there are two functions for loading data generated in the DICOMM format created by the Humphrey Field Analyzer (HFA) by Zeiss. Theoretically, this function should also load data from other perimeters.

@@

The VisualField package provided normative data for the Humphrey Field Analyzer but currently not for other perimeters. The loadoctopus function also imports some normative data from the Octopus perimeter by Haag-Streit, but this does not allow full use of the functionality of the visualFields package, currently.

>>>>

# What is visual field testing?

Visual field testing is a psychophysical test that measures a patient's visual ability at different locations of the visual field. This ability is usually highest in the center of the visual field (i.e. the point that the observer looks at) and decreases toward the periphery (hill of vision).

@@

This dataset contains the results of 42 visual field tests of one patient (id: sample1). The patient has a chronic-neurodegenerative disease of the optic nerve called primary open-angle glaucoma (type: pwg).

@@

## Plotting the visual field

# Visual field tests

For the typical visual field test, the subject or patient is comfortably seated in front of a perimeter and puts his head on a chin rest looking into a white Ganzfeld cupule (or hemisphere). The perimeter is located in a dimly lit, calm room and the subject has been explained what to do by an experienced technician before the examination. In his hand, the subject has a button in order to signal whether he saw a stimulus.

@@

The standard environment assumes a static white-on-white perimetry with a Humphrey perimeter, using a 24-2 grid, a SITA standard algorithm, a stimulus size of Goldman Size III. The stimulus duration and the $L_{max}$ are not explicitly stated in the info.

@@

The spatial position of the points \$l_1 \$to $ln$ in the visual field is found in the locmap and depends on the pattern used by the perimeter. When analysing data, you need to specify the locmap.

@@

Note that some grids spare the blind spot, either on purpose or because they cover only locations central to the blind spot. In this case, locmap\$bs is numeric(0).

@@

# Measurements

Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response. It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.

@@

# Differential light sensitivity and other mathematical derivatives of threshold

However, the data analyzed with the visual fields package usually consists of the subjects *differential light sensitivity* for a stimulus at a given location in decibel (dB) as the most elementary unit. This sensitivity is calculated from the psychophysical threshold, which is derived from the aforementioned responses at this location.

@@

## False negatives / positives

Residual deviation from the psychometric function may occur due to inattention of the observer but also due to short-term fluctuation of the threshold. For example, it has been shown that the threshold underlies short-term fluctuates in the areas where light sensitivity was negatively affected by glaucoma, a disease of the optic nerve.

@@

The resulting difference is also measured in dB and is called either *deviation* (loss of function has a negative sign; used in Humphrey perimeters) or *defect* (loss of function has a positive sign; used in Octopus perimeters). In this package, *deviation* is used, and it is called *total deviation*. Make sure that you get the right sign if you import data.

@@

### Dataset

We use the dataset included in the package for demonstrating calculation of global parameters.

@@

vfpwgSunyiu24d2[1:4, 1:12]

@@

The standard already contains some additional information with a border presenting a color-coded visualization of the deviation and a graphical representation of the blind spot. Note that the vfplot with the option "td" needs information on differential light-sensitivities in a normal reference population and on the probability of given deviations from these normal values. These informations are included for the typical Humphrey Field Analyzer fields, but you need to collect this information if you use other parameters.

@@

@@

vfplot(example1, type = "tds")

vfplot(example1, type = "pd")

vfplot(example1, type = "pds")

@@

Again, the probability of the observed values given that the visual field is normal can be calculated.

@@

Monitoring function over time in glaucoma patients is the most common clinical application of visual fields in ophthalmology.

For several reasons, measurements will fluctuate over time. Fluctuation between measurements will be higher in eyes with glaucoma, especially in areas of glaucoma-related visual field defects. Therefore, it is not possible to compare two field tests and claim that the disease has progressed only because the second test yields poorer sensitivities.

Thus, statistical methods and a minimum number of tests are necessary in order to demonstrate disease progression in the visual fields. Statistical methods that can be subdivided into *event-based* and *trend-based* methods. Furthermore, methods can be subdivided into those that require datasets which describe the "normal fluctuation" in a given population, those that rely on a mathematical model of visual field progression, and those that do not rely on either.

Event-based methods are not implemented in the visualFields pacakge. Briefly, they look for progression between the most recent field and a baseline test. Progression is characterized by a decrease in sensitivity with a magnitude that is unlikely (< 5%) in the population under examination (population based data is necessary) and needs to be confirmed in further tests (usually at least two).

Trend-based methods assume that progression takes place more slowly, so that several field tests are carried out during this time span. This concept is unsettling for the clinician who wants to stop progression the first time that he sees the patient, but regularly takes place even in the case of optimal treatment. The advantage is that all measurements are taken into account, not only the baseline test and the most recent tests.

## Dataset

For the next example, we will select a series of 27 visual fields that were obtained over several years.

@@

## Plotting progression

The sparkline plot only shows the raw data.

@@

vfsparklines(example2)

@@

The lego plot shows probabilities for progression which relies on population based data.

vflegoplot(example2)

## Statistical analysis

@@

The *glr* function performs linear regression on global parameters.

@@

The *plr* function performs pointwise linear regression.

prog1 <- plr(example2)
str(prog1, max.level = 1)

print(sum(prog1$pval < 0.05))

@@

The *poplr* function performs pointwise linear regression, and then calculates an S-score on the visual field series and on random permutations of the series, in order to control the Type I error (diagnosing progression in a non-progressing series) without relying on population data or a mathematical model of progression. It is computationally intensive.

The *poplr* function yields a large list with many intermediate computations. The main result is the S-score and the probability of getting this S-score in the absence of progression (Type I error). Note that there is a left and a right probability (deterioration and improvement) in the current implementation.

@@

prog2 <- poplr(example2)
str(prog2, max.level = 1)

@@

An improvement is highly unlikely in this series.

prog2$csr
prog2$csrp

In contrast, a deterioration is very likely (p < 0.0001).

prog2$csl
prog2$cslp

## Creating pretty reports

@@

```{r eval=FALSE, include=TRUE}

vfspa(example2, file = "test.pdf")

vfspashiny(example2, file = "test.pdf")

@@

DICOM (Digital Imaging and Communications in Medicine) standards for ophthalmic static perimetry are specified in Supplement 146. However, there are two functions for loading data generated in the DICOMM format created by the Humphrey Field Analyzer (HFA) by Zeiss.

Currently, the data representation in the VisualFields package differs from these standards and data cannot be exported from the package to DICOMM.

@@

Humphrey Field Analyzer (Carl-Zeiss-Meditec)

The VisualField package provided normative data for the Humphrey Field Analyzer but currently not for other perimeters.

Octopus 900 Perimeter (Haag-Streit)

The loadoctopus function also imports some normative data from the Octopus perimeter by Haag-Streit, but this does not allow full use of the functionality of the visualFields package, currently.

For making full use of the visusalField package with other parameters, you ideally need a database of visual fields from a normal reference population, but the creation of normal values with the visualFields package goes beyond the scope of this vignette.
